A New Method of Modelling Tuneable Lasers with Functional Composition

### 2020-02-06 Papers/Journal/Tuneable_Lasers.tex
Fixed spacing in derivatives
Added introduction on wave breaking and modulation instability

\providecommand{\df}{\textrm{d}}
\newcommand{\diff}[3][]{\frac{\textrm{d}^{#1}#2}{\textrm{d}{#3}^{#1}}}
\newcommand{\pdiff}[3][]{\frac{\partial^{#1}#2}{\partial{#3}^{#1}}}
\newcommand{\Es}{E_{\textrm{sat}}}
\newcommand{\FT}[1]{\mathcal{F}\left\{ #1 \right\}}
\newcommand{\FTi}[1]{\mathcal{F}^{-1}\left\{ #1 \right\}}
\newcommand{\Her}[2]{\widetilde{H}_{#1} \left( #2 \right)}
%\newcommand{\eps}{\varepsilon}
%\newcommand{\rect}[1]{\textrm{rect}\left( #1 \right)}

@@

\section{Introduction}
\label{sec:intro}
A tuneable laser has the ability to vary the frequency of its output by up to about 100 nanometres \cite{bohun, burgoyne2010, yamashita}. Tuneable lasers simultaneously lase at all frequencies within this bandwidth. This tuneability leads to applications in spectroscopy and high resolution imaging, such as coherent anti-Stokes Raman spectroscopy and optical coherence tomography \cite{bohun, burgoyne2014, yamashita}, as well as communications and diagnostics of ultrafast processes \cite{silfvast}. A typical tuneable ring laser cavity is shown in Figure \ref{fig:cavity}.

\subsection{Wave Breaking}

\subsection{Modelling Efforts}
\label{sec:modelling}
The standard equation for studying nonlinear optics is the nonlinear Schr\"odinger equation (NLSE),
\begin{align}
	\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + i \gamma |A|^2 A.
	\label{eq:smallnlse}
\end{align}
Here $A = A(T, z) : \mathbb{R}^2 \mapsto \mathbb{C}$ is the complex pulse amplitude, $\beta_2 \in \mathbb{R}$ is the second order dispersion, and $\gamma \in \mathbb{R}$ is the coefficient of nonlinearity. In practice, the NLSE, \eqref{eq:smallnlse}, lacks a few key terms, thus, it is often generalized by adding amplification, loss, and occasionally higher order terms. These additions give the generalized nonlinear Schr\"{o}dinger equation (GNLSE) \cite{agrawal2013, bohun, finot, peng, shtyrina, yarutkina},
	\begin{align}
	\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + i \gamma |A|^2 A + \frac{1}{2}g(A) A - \alpha A,
	\label{eq:nlse}
\end{align}
where $g(A)$ is an amplifying term due to the gain, and $\alpha \in \mathbb{R}^+$ is the loss due to scattering and absorption.

The GNLSE has many applications in nonlinear optics and fibre optic communications, however, in laser physics typically a modulation term is also added to ensure mode-locking, this yields the master equation of mode-locking \cite{hausbook, haus1975, haus1986, haus1992, haus2000, tamura1996, usechak},
\begin{align}
	\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + i \gamma |A|^2 A + \frac{1}{2}g(A) A - \alpha A - M(T).
	\label{eq:meml}
\end{align}
No analytic solution is known for \eqref{eq:meml}---even after making the assumptions that the gain is constant, and the modulation is quadratic \cite{hausbook, haus1975, haus1996}. However, by neglecting certain terms of \eqref{eq:meml} analytic solutions can be found \cite{burgoyne2014, haus1975, haus1986, haus1991, haus1992, haus1996, tamura1996, usechak}. Furthermore, see Haus \cite{haus2000} for a comprehensive description and history of the theory of mode-locked lasers.

\begin{figure}[tbp]
	\centering
	\input{Cavity}
	\caption{Typical cavity of a fibre based tuneable laser. The laser pulses travel clockwise around each loop.}
	\label{fig:cavity}
\end{figure}

@@

Such a method was first proposed in 1955 by Cutler \cite{cutler} while analyzing a microwave regenerative pulse generator. This method was adapted for mode-locked lasers in 1969 by Siegman and Kuizenga \cite{kuizenga1970a, kuizenga1970b, kuizenga1970, siegman}. The effects of the nonlinearity would not be considered until Martinez, Fork, and Gordon \cite{martinez1984, martinez1985} tried modelling passively mode-locked lasers. This issue has recently been readdressed by Burgoyne \cite{burgoyne2014} in the literature for tuneable lasers. In each of these models the effect of each block is described by a transfer function.

@@

To proceed we must choose the form of the amplification term, $g(A)$. We assume the gain has the form
\begin{align}
	g(A) &= \frac{g_0}{1 + E / \Es},& E &= \int_{-\infty}^\infty |A|^2 \, \df T,
	\label{eq:energy}
\end{align}
where $g_0$ is a small signal gain, $E$ is the energy of the pulse, and $\Es$ is the energy at which the gain begins to saturate \cite{bohun, burgoyne2014, haus1975, hausbook, haus1992, haus2000, haus1991, kartner, peng, shtyrina, silfvast, usechak, yarutkina}. This leads to
\begin{align}
	G(A) &= \left( \frac{\Es}{E} W \left( \frac{E}{\Es} \textrm{e}^{E/\Es} \textrm{e}^{g_0 L_g} \right) \right)^{1/2} A
\end{align}
for the amplification of the pulse within the Er-doped fibre. Continuing this process for the loss, dispersion, and nonlinearity yields
\begin{align}
	L(A) &= (1 - R) \textrm{e}^{- \alpha L_T}A, \\
	D(A) &= \FTi{\textrm{e}^{i \omega^2 L_D\beta_2/2} \FT{A}}, \\
	F(A) &= A \textrm{e}^{i \gamma |A|^2 L_f},
\end{align}
where $R$ is the reflectivity of the output coupler\footnote{Depending on the layout of the laser cavity the loss may instead take the form $L(A) = R \textrm{e}^{- \alpha L_T}A$.}, $L_T$ is the total length of the laser circuit, $L_D$ is the length of the dispersive medium, $L_f$ is the length of fibre between the amplifier and the output coupler, and where $\mathcal{F}$ denotes the Fourier transform. Finally, we must assume a form for the modulation. The modulation is considered to be applied externally in which ever way the operator sees fit. For simplicity the representation is taken as the Gaussian
\begin{align}
	M(A) &= \textrm{e}^{-T^2 / 2 T_M^2} A,
\end{align}
where $T_M$ is the characteristic width of the modulation.

\subsection{Non-Dimensionalization}
The structure of each process of the laser can be better understood by re-scaling the time, energy, and amplitude; this suggests the convenient scalings:
\begin{align}
	T &= T_M \widetilde{T},& E &= \Es \widetilde{E},& A &= \left( \frac{\Es}{T_M} \right)^{1/2} \widetilde{A}.
\end{align}
Revisiting each process map shows each process has a characteristic non-dimensional parameter. The new mappings---after dropping the tildes---are
\begin{equation}
	\begin{aligned}
		G(A) &= \left(E^{-1} W \left( a E \textrm{e}^{E}\right) \right)^{1/2} A, & F(A) &= A \textrm{e}^{i b |A|^2}, \\
		D(A) &= \FTi{\textrm{e}^{i s^2 \omega^2} \FT{A}}, & L(A) &= h A, \\
		M(A) &= \textrm{e}^{-T^2 / 2} A,
		\label{eq:effects}
	\end{aligned}
\end{equation}
with the four dimensionless parameters,
\begin{equation}
	\begin{aligned}
		a &= \textrm{e}^{g_0 L_g} \sim 8 \times 10^3,& \qquad h &= (1 - R) \textrm{e}^{-\alpha L} \sim 0.04, \\
		b &= \gamma L_f \frac{\Es}{T_M} \sim 1,& \qquad s &= \sqrt{\frac{\beta_2 L_D}{2 T_M^2}} \sim 0.2,
		\label{eq:ndparam}
	\end{aligned}
\end{equation}
which characterize the behaviour of the laser. Nominal values for the parameters, \eqref{eq:ndparam}, can be found in \cite{agrawal2002, agrawal2013, bohun, burgoyne2014, burgoyneemail, finot, li, litchinitser, peng, shtyrina, tamura1993, tamura1996, tomlinson, usechak, yamashita, yarutkina}. Notice that the modulation is only characterized by $T_M$, and each other process has its own associated independent non-dimensional parameter.

@@

\section{Results}

\subsection{Linear Solution}

\subsection{Nonlinear Solution and Wave Breaking}

@@

\section{Linear Solution}
By neglecting the effoct of the nonlinearity, that is, $b = 0$, a solution can be found analytically. We assume the solution will have the form of a chirped Gaussian. There are a few reasons for this; the solution to the models presented in \cite{cutler, siegman, kuizenga1970a, martinez1984, martinez1985} were Gaussian, the equilibrium shape will be highly correlated to the shape of the modulation function, and since a Gaussian is a fixed point of the Fourier transform \cite{gradshteyn}.

To compute $\mathcal{L}(A)$, consider the pulse
\begin{align}
	A = \sqrt{P} \exp \left( -(1 + iC) \frac{T^2}{2 \sigma^2} \right) \textrm{e}^{i \phi_0},
	\label{eq:A0}
\end{align}
where $P$ is the peak power, $C$ is the chirp, $\sigma^2$ is the variance, and $\phi_0$ is the initial phase. This form is chosen because it resembles the envelope and linear chirping found by experiment ***cite.

@@

\begin{figure}[htbp]
	\centering
	\includegraphics{ParamSpaceErr}
	\caption{}
	\label{fig:error}
\end{figure}

\begin{figure}[htbp]
	\centering
	\includegraphics{ParamSpaceEnergy}
	\caption{}
	\label{fig:}
\end{figure}

>>>>

\providecommand{\df}{\textrm{d}}
\newcommand{\diff}[3][\hspace{-0.5pt}]{\frac{\textrm{d}^{#1}#2}{\textrm{d}{#3}^{#1}}}
\newcommand{\pdiff}[3][\hspace{-0.5pt}]{\frac{\partial^{#1}#2}{\partial{#3}^{#1}}}
\newcommand{\Es}{E_{\textrm{sat}}}
\newcommand{\FT}[1]{\mathcal{F}\left\{ #1 \right\}}
\newcommand{\FTi}[1]{\mathcal{F}^{-1}\left\{ #1 \right\}}
\newcommand{\Her}[2]{\widetilde{H}_{#1} \left( #2 \right)}
\DeclareMathOperator{\sech}{sech}
%\newcommand{\eps}{\varepsilon}
%\newcommand{\rect}[1]{\textrm{rect}\left( #1 \right)}

@@

\section{Introduction}
\label{sec:intro}
A tuneable laser has the ability to vary the frequency of its output by up to about 100 nanometres \cite{bohun2015, burgoyne2010, yamashita2009}. Tuneable lasers simultaneously lase at all frequencies within this bandwidth. This tuneability leads to applications such as optical coherence tomography \cite{bohun2015, burgoyne2014, yamashita2009}, coherent anti-Stokes Raman spectroscopy \cite{burgoyne2014}, deep tissue multi-photon microscopy \cite{chung2017, liu2017}, and diagnostics of ultrafast processes \cite{burgoyne2014, silfvast2004}. A typical tuneable ring laser cavity is shown in Figure \ref{fig:cavity}.

The nonlinearity causes several effects to arise within the laser cavity due to the interplay of dispersion, modulation, and the nonlinearity \cite{bohun2015, coen1997, lapre2019, shao2019, woodward2018}. The effect of most interest for this paper is optical wave breaking which causes the leading edge of a pulse to be red shifted, and the trailing edge blue shifted\footnote{In a normally dispersive optical fibre.} to the point that a shock is developed \cite{anderson1992, rothenberg1989a, rothenberg1989b, tomlinson1984, tomlinson1985}. The frequency shifts cause the pulse to become rectangular in the frequency domain with a linear chirp over most of the pulse. Wave breaking manifests from self-phase modulation (SPM) which is a direct effect of the nonlinearity in which the pulse interferes with itself. This interference generates new frequencies which induces the rectangular profile \cite{agrawal2013, mollenauer1980, woodward2018}. Modulation instablility is another effect that will be of interest that typically arises in the anomolous dispersion regime. However, in the presence of multiple pulses modulation instability can emerge in the normal dispersion regime as well through cross-phase modulation (XPM) and four-wave mixing (FWM) \cite{agrawal1987, agrawal1989, agrawal2013, haelterman1992}. Modulation instability causes a wave to break into short pulses, and can cause a laser to lose mode-locking \cite{agrawal1987, coen1997, haelterman1992}. Within a ring laser cavity wave breaking and modulation instability can become parasitic leading to an unstable and unsustainable pulse. This gives rise to the need of understanding the rich landscape of the parameter space, and determining design principles to guarantee the ring laser is stable and sustainable \cite{bohun2015, burgoyneemail, finot2008, lapre2019, woodward2018}.

\section{Modelling Efforts}
\label{sec:modelling}
The standard equation for studying nonlinear optics is the nonlinear Schr\"odinger equation (NLSE),
\begin{align}
	\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + i \gamma |A|^2 A.
	\label{eq:smallnlse}
\end{align}
Here $A = A(T, z) : \mathbb{R}^2 \mapsto \mathbb{C}$ is the complex pulse amplitude, $\beta_2 \in \mathbb{R}$ is the second order dispersion, and $\gamma \in \mathbb{R}$ is the coefficient of nonlinearity. In practice, the NLSE, \eqref{eq:smallnlse}, lacks a few key terms, thus, it is often generalized by adding amplification and loss\footnote{Occasionally higher order terms are added as well.}. These additions give the generalized nonlinear Schr\"{o}dinger equation (GNLSE) \cite{agrawal2013, bohun2015, finot2008, peng2018, shtyrina2017, yarutkina2013},
	\begin{align}
	\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + i \gamma |A|^2 A + \frac{1}{2}g(A) A - \alpha A,
	\label{eq:nlse}
\end{align}
where $g(A)$ is an amplifying term due to the gain, and $\alpha \in \mathbb{R}^+$ is the loss due to scattering and absorption.

\begin{figure}[tbp]
	\centering
	\input{Cavity}
	\caption{Typical cavity of a fibre ring tuneable laser \cite{burgoyne2014, chung2017, lapre2019, shao2019, tang2014}. The laser pulses travel clockwise around each loop.}
	\label{fig:cavity}
\end{figure}

The GNLSE has many applications in nonlinear optics and fibre optic communications, however, in laser physics typically a modulation term is also added to ensure mode-locking, this yields the master equation of mode-locking \cite{haus1984, haus1975, haus1986, haus1992, haus2000, tamura1996, usechak2005},
\begin{align}
	\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + i \gamma |A|^2 A + \frac{1}{2}g(A) A - \alpha A - M(T).
	\label{eq:meml}
\end{align}
No analytic solution is known for \eqref{eq:meml}---even after making the assumptions that the gain is constant, and the modulation is quadratic \cite{haus1984, haus1975, haus1996}. However, by neglecting certain terms of \eqref{eq:meml} analytic solutions can be found \cite{burgoyne2014, haus1975, haus1986, haus1991, haus1992, haus1996, tamura1996, usechak2005}. Furthermore, see Haus \cite{haus2000} for a comprehensive description and history of the theory of mode-locked lasers.

@@

Such a method was first proposed in 1955 by Cutler \cite{cutler1955} while analyzing a microwave regenerative pulse generator. This method was adapted for mode-locked lasers in 1969 by Siegman and Kuizenga \cite{kuizenga1970a, kuizenga1970b, kuizenga1970, siegman1969}. The effects of the nonlinearity would not be considered until Martinez, Fork, and Gordon \cite{martinez1984, martinez1985} tried modelling passively mode-locked lasers. This issue has recently been readdressed by Burgoyne \cite{burgoyne2014} in the literature for tuneable lasers. In each of these models the effect of each block is described by a transfer function.

@@

To proceed we must choose the form of the amplification term, $g(A)$. We assume the gain has the form
\begin{align}
	g(A) &= \frac{g_0}{1 + E / \Es},& E &= \int_{-\infty}^\infty |A|^2 \, \df T,
	\label{eq:energy}
\end{align}
where $g_0$ is a small signal gain, $E$ is the energy of the pulse, and $\Es$ is the energy at which the gain begins to saturate \cite{bohun2015, burgoyne2014, haus1975, haus1984, haus1992, haus2000, haus1991, kartner2005, peng2018, shtyrina2017, silfvast2004, usechak2005, yarutkina2013}. This leads to
\begin{align}
	G(A) &= \left( \frac{\Es}{E} W \left( \frac{E}{\Es} \textrm{e}^{E/\Es} \textrm{e}^{g_0 L_g} \right) \right)^{1/2} A
\end{align}
for the amplification of the pulse within the Er-doped fibre. Continuing this process for the loss, dispersion, and nonlinearity yields
\begin{align}
	L(A) &= (1 - R) \textrm{e}^{- \alpha L_T}A, \\
	D(A) &= \FTi{\textrm{e}^{i \omega^2 L_D\beta_2/2} \FT{A}}, \\
	F(A) &= A \textrm{e}^{i \gamma |A|^2 L_f},
\end{align}
where $R$ is the reflectivity of the output coupler\footnote{Depending on the layout of the laser cavity the loss may instead take the form $L(A) = R \textrm{e}^{- \alpha L_T}A$.}, $L_T$ is the total length of the laser circuit, $L_D$ is the length of the dispersive medium, $L_f$ is the length of fibre between the amplifier and the output coupler, and where $\mathcal{F}$ denotes the Fourier transform. Finally, we must assume a form for the modulation. The modulation is considered to be applied externally in which ever way the operator sees fit. For simplicity the representation is taken as the Gaussian
\begin{align}
	M(A) &= \textrm{e}^{-T^2 / 2 T_M^2} A,
	\label{eq:modform}
\end{align}
where $T_M$ is the characteristic width of the modulation. This assumption is not particularly restrictive. Calcaterra and Boldt have shown that Gaussians can form a basis for $L^2(\mathbb{R})$ \cite{calcaterra2008a}. Therefore, we can approximate any modulation function in $L^2(\mathbb{R})$ by a sum of \eqref{eq:modform}.

\subsection{Non-Dimensionalization}
The structure of each process of the laser can be better understood by re-scaling the time, energy, and amplitude; this suggests the convenient scalings:
\begin{align}
	T &= T_M \widetilde{T},& E &= \Es \widetilde{E},& A &= \left( \frac{\Es}{T_M} \right)^{1/2} \widetilde{A}.
\end{align}
Revisiting each process map shows each process has a characteristic non-dimensional parameter. The new mappings---after dropping the tildes---are
\begin{equation}
	\begin{aligned}
		G(A) &= \left(E^{-1} W \left( a E \textrm{e}^{E}\right) \right)^{1/2} A, & F(A) &= A \textrm{e}^{i b |A|^2}, \\
		D(A) &= \FTi{\textrm{e}^{i s^2 \omega^2} \FT{A}}, & L(A) &= h A, \\
		M(A) &= \textrm{e}^{-T^2 / 2} A,
		\label{eq:effects}
	\end{aligned}
\end{equation}
with the four dimensionless parameters,
\begin{equation}
	\begin{aligned}
		a &= \textrm{e}^{g_0 L_g} \sim 8 \times 10^3,& \qquad h &= (1 - R) \textrm{e}^{-\alpha L} \sim 0.04, \\
		b &= \gamma L_f \frac{\Es}{T_M} \sim 1,& \qquad s &= \sqrt{\frac{\beta_2 L_D}{2 T_M^2}} \sim 0.2,
		\label{eq:ndparam}
	\end{aligned}
\end{equation}
which characterize the behaviour of the laser. Nominal values for the parameters, \eqref{eq:ndparam}, can be found in \cite{agrawal2002, agrawal2013, bohun2015, burgoyne2014, burgoyneemail, finot2008, li1998, litchinitser1997, peng2018, shtyrina2017, tamura1993, tamura1996, tomlinson1984, usechak2005, yamashita2009, yarutkina2013}. Notice that the modulation is only characterized by $T_M$, and each other process has its own associated independent non-dimensional parameter.

@@

\section{Results}
We split the results into two subsections. In the following subsection we investigate the low nonlinearity limit, and in Section \ref{sec:nlresults} we look at the full nonlinear model.

\subsection{Linear Solution}
By neglecting the effect of the nonlinearity, that is, $b = 0$, a solution can be found analytically. We assume the solution will take the form of a chirped Gaussian. There are a few reasons for this; the solution to the models presented in \cite{cutler1955, siegman1969, kuizenga1970a, martinez1984, martinez1985} were Gaussian, the equilibrium shape will be highly correlated to the shape of the modulation function, and since a Gaussian is a fixed point of the Fourier transform \cite{gradshteyn2007}. Furthermore, this form is chosen because it resembles the envelope and linear chirp expected \cite{burgoyne2014, haus1975, haus1996, haus2000, usechak2005}.

To compute $\mathcal{L}(A)$, consider the pulse
\begin{align}
	A = \sqrt{P} \exp \left( -(1 + iC) \frac{T^2}{2 \sigma^2} \right) \textrm{e}^{i \phi_0},
	\label{eq:A0}
\end{align}
where $P$ is the peak power, $C$ is the chirp, $\sigma^2$ is the variance, and $\phi_0$ is the initial phase. This pulse shape is depicted in Figure \ref{fig:samplegauss}.

\begin{figure}[tbp]
	\includegraphics{Sample_Gauss}
	\caption{Solution to the linear model given by \eqref{eq:A0}.}
	\label{fig:samplegauss}
\end{figure}

Our linearized model indeed exhibits the same form as previous linear models. However, by including the effects of the nonlinearity we uncover the rich interplay between dispersion, modulation, and nonlinearity not previously found mathematically.

\subsection{Nonlinear Solution and Instability}
\label{sec:nlresults}

In this iterative scheme---as well as within the laboratory---we must specify the input pulse shape. The most common form is a hyperbolic secant \cite{coen1997, finot2008, mitschke1986, rothenberg1989b, tomlinson1984}, which we assume has the exact form
\begin{align}
	A_0 = \Gamma \sech \left( 2 T \right) \textrm{e}^{i \pi / 4},
\end{align}
where $\Gamma$ is a normalizing factor chosen so the pulse has an initial energy of $E_0 = 0.1$.

xpm \cite{agrawal2013, agrawal1987, agrawal1989}

\begin{figure}[tbp]
	\centering
	\includegraphics{ParamSpaceErr}
	\caption{}
	\label{fig:error}
\end{figure}

\begin{figure}[tbp]
	\centering
	\includegraphics{ParamSpaceEnergy}
	\caption{}
	\label{fig:}
\end{figure}

\begin{figure}[tbp]
	\centering
	\includegraphics{ParamSpaceEnergySwitch}
	\caption{}
	\label{fig:energyswitch}
\end{figure}

@@

@@
